Experimental and numerical studies of the effects of resonant and random excitations of the proton beam in the LHC, with applications to the design of a pulsed hollow electron lens for active halo control

### 2018-03-22 resex.tex
First complete revision of bibliography

\documentclass[
prstab,reprint
%,draft
%,preprint,tightenlines
,linenumbers
,longbibliography
,preprintnumbers
,showpacs
,showkeys
,byrevtex
,amsfonts,amssymb,amsmath
,floatfix
%,endfloats
]{revtex4-1}

@@

\begin{figure*}
  \begin{tabular}{cc}
    \includegraphics[width=0.4\textwidth]{CHG1b_170512_8p75A_2-4-2p7kG_500V_76mA_hires_Emap} &
    \includegraphics[width=0.4\textwidth]{CHG1b_170512_8p75A_2-4-2p7kG_500V_76mA_hires_Eslice}\\
  \end{tabular}
  \caption{Calculated relative electric field for the hollow electron
    gun CHG1b in the transverse $x$-$y$ plane (left) and as
    1-dimensional cuts through the $x$ and $y$ axes (right). The field
    calculations are based on measurements at the Fermilab
    electron-lens test stand combined with \code{Warp} calculations of
    the electric potentials and fields in a cylindrical beam pipe.}
  \label{core:fig:1}
\end{figure*}

@@

To estimate the dipole component of the kicks that originate from the
injection and extraction bends, we used the approach described in
Ref.~\cite{hel_bends_stancari}. The bends are modeled as a bent
cylindrical pipe with a static charge distribution of electrons. The
resulting electric field is calculated, with the vacuum pipe as
boundary, using the solvers of the \code{Warp} particle-in-cell
code~\cite{warp, Friedman:IEEE-plasma:2014}. (The contribution of the
magnetic field generated by the electron current was neglected in this
study.) The field, integrated over the trajectory of the protons, is
then translated into a symplectic kick map.

@@

For a given HL-LHC HEL configuration, the corresponding
current-density profile measured in the Fermilab test stand is used as
input to calculate the electromagnetic fields. For the purpose of
estimating the residual fields, the measured distribution is
compressed to the inner electron beam radius of 1.24~mm, corresponding
to $4\sigma_p$, as described in Table~\ref{tab:hllhc_param} and
Table~\ref{tab:hel_param}. A distribution with about $65\,000$
particles is generated according to the measured current-density
profile. As boundary condition, the LHC inner beam pipe radius
$b = \q{30}{mm}$ is used. The potential and fields are then calculated
with \code{Warp}~\cite{warp, Friedman:IEEE-plasma:2014}. The resulting
relative electric field strengths are shown in
Fig.~\ref{core:fig:1}. The electric field is obviously proportional to
the charge density of electrons. Its relative strength is rather
insensitive to the hollow beam radii, as long as these remain small
compared to the beam pipe radius. Further details on the measurements
and \code{Warp} simulations can be found in
Ref.~\cite{hel_res_field_stancari_2017}.

@@

\begin{table*}
  \caption{Summary of parameters used in numerical simulations of
    distribution tracking and frequency-map analysis (FMA). Further
    details can be found in Refs.~\cite{md_sim_hel_res_ex_fitterer,
      resexmd2017}.}
  \label{tab:sim_param}
  \begin{ruledtabular}
    \begin{tabular}{lcc}
      Parameter & distribution & FMA \\
      \colrule
      beam &\multicolumn{2}{c}{Beam~1} \\
      beam energy &\multicolumn{2}{c}{450 GeV} \\
      normalized emittance & \q{3.5}{\mu m} & \q{2.5}{\mu m} \\
      $4\sigma$ bunch length & \multicolumn{2}{c}{1.3~ns} \\
      $1\sigma$ bunch length & \multicolumn{2}{c}{9.7~cm} \\
      particle distribution & \parbox{0.4\textwidth}{6D Gaussian
                              distribution with $10^4$ particles}
                               & \parbox{0.4\textwidth}{equally spaced
                                 grid in $x, y$ up to $10\sigma$, with
                                 $(\Delta p / p_0) = 0$} \\
      turns tracked & $10^6$ & $10^4$ \\
      \hline
      optics & \multicolumn{2}{c}{2016 or 2017 injection optics, with
               $\beta^* = \q{11}{m}$ at IP1 and IP5} \\
      machine imperfections & standard errors, with $a_1 = b_1 =
                              0$\footnote{Orbit
                              errors are disabled due to
                              different implementation of the
                              $a_1, b_1$ coefficients in \code{Lifetrac} and
                              \code{MAD-X}. $b_2$ errors are adjusted to
                              yield an average
                              peak $\beta$-beat of 15\% over 60~seeds, as
                              observed in optics measurements
                              in the LHC.} & no errors \\
      octupoles  & \multicolumn{2}{c}{$I_{\mathrm{MOF}} =
                   \q{+19.6}{A}$, $I_{\mathrm{MOD}} = \q{-19.6}{A}$} \\
      tunes $(Q_x, Q_y)$ & \multicolumn{2}{c}{(64.28, 59.31) for 2016,
                           (62.27, 60.295) for 2017} \\
      chromaticities $(Q'_x, Q'_y)$ & \multicolumn{2}{c}{(+15, +15)} \\
      \hline
      transverse aperture & $5.7\sigma$ &  \\
      longitudinal aperture & $10\sigma$ &  \\
    \end{tabular}
  \end{ruledtabular}
\end{table*}

For the preparation and interpretation of the experiments, two
different types of simulations were performed:
%
\begin{itemize}
\item tracking of a Gaussian distribution, referred to as
  `distribution tracking' in this paper, to obtain particle loss rates
  and emittance evolution;
\item frequency-map analysis (FMA), to visualize the location and
  intensity of resonances~\cite{fmalaskar}.
\end{itemize}
%
For both simulation types, the tracking code
\code{Lifetrac}~\cite{lifetrac} was used. The simulation parameters
are summarized in Table~\ref{tab:sim_param}. Further details are given
in Refs.~\cite{md_sim_hel_res_ex_fitterer, resexmd2017}.

@@

The random excitation is a powerful and flexible halo-cleaning
pattern, but it is also potentially harmful for the proton beam
core. The effects of random noise in various machines have been
studied extensively in the past (see, for instance,
Refs.~\cite{noise_lebedev_ssc, noise_alexahin_lhc, noise_2007_ohmi,
  noise_2014_ohmi, md400_noise_injection}).

@@

\begin{acknowledgments}
  There are several people we would like to acknowledge for their
  support, insights, participation in studies, and assistance in data
  collection. In particular, Giulia Pa\-potti offered essential
  support for machine operations and helped prepare the 2016
  experiments. Roderik Bruce and Gianluca Va\-len\-ti\-no made
  important suggestions on the design of the experiments. Dmitry
  Sha\-ti\-lov shared his deep understanding of numerical tracking
  simulations with the \code{Lifetrac} code. St\'{e}\-phane
  Far\-toukh, Ric\-car\-do De Maria and Ro\-ge\-lio Tom\'{a}s offered
  invaluable assistance in generating the appropiate accelerator
  lattice models. We wish to thank Sergei Na\-gai\-tsev for the
  insightful discussions on the interaction between the transverse
  damping system and resonant excitations. We are grateful to
  En\-ri\-co Bra\-vin, Georges Trad, and the whole beam
  instrumentation team for their crucial support on
  synchrotron-radiation diagnostics. We thank Ste\-fania
  Pa\-pa\-do\-pou\-lou and Fa\-nou\-ria An\-to\-niou for the fruitful
  collaboration on the analysis of the synchrotron-radiation profiles.

>>>>

\documentclass[
prstab,reprint
%,draft
%,preprint,tightenlines
,linenumbers
,longbibliography
,preprintnumbers
,showpacs
,showkeys
,amsfonts,amssymb,amsmath
,floatfix
]{revtex4-1}

@@

\begin{figure*}
  \begin{tabular}{cc}
    \includegraphics[width=0.4\textwidth]{CHG1b_170512_8p75A_2-4-2p7kG_500V_76mA_hires_Emap} &
    \includegraphics[width=0.4\textwidth]{CHG1b_170512_8p75A_2-4-2p7kG_500V_76mA_hires_Eslice}\\
  \end{tabular}
  \caption{Calculated relative electric field for the hollow electron
    gun CHG1b in the transverse $x$-$y$ plane (left) and as
    1-dimensional cuts through the $x$ and $y$ axes (right). The field
    calculations are based on measurements at the Fermilab
    electron-lens test stand combined with \code{warp} calculations of
    the electric potentials and fields in a cylindrical beam pipe.}
  \label{core:fig:1}
\end{figure*}

@@

To estimate the dipole component of the kicks that originate from the
injection and extraction bends, we used the approach described in
Ref.~\cite{hel_bends_stancari}. The bends are modeled as a bent
cylindrical pipe with a static charge distribution of electrons. The
resulting electric field is calculated, with the vacuum pipe as
boundary, using the solvers of the \code{warp} particle-in-cell
code~\cite{warp, Friedman:IEEE-plasma:2014}. (The contribution of the
magnetic field generated by the electron current was neglected in this
study.) The field, integrated over the trajectory of the protons, is
then translated into a symplectic kick map.

@@

For a given HL-LHC HEL configuration, the corresponding
current-density profile measured in the Fermilab test stand is used as
input to calculate the electromagnetic fields. For the purpose of
estimating the residual fields, the measured distribution is
compressed to the inner electron beam radius of 1.24~mm, corresponding
to $4\sigma_p$, as described in Table~\ref{tab:hllhc_param} and
Table~\ref{tab:hel_param}. A distribution with about $65\,000$
particles is generated according to the measured current-density
profile. As boundary condition, the LHC inner beam pipe radius
$b = \q{30}{mm}$ is used. The potential and fields are then calculated
with \code{warp}~\cite{warp, Friedman:IEEE-plasma:2014}. The resulting
relative electric field strengths are shown in
Fig.~\ref{core:fig:1}. The electric field is obviously proportional to
the charge density of electrons. Its relative strength is rather
insensitive to the hollow beam radii, as long as these remain small
compared to the beam pipe radius. Further details on the measurements
and \code{warp} simulations can be found in
Ref.~\cite{hel_res_field_stancari_2017}.

@@

\begin{table*}
  \caption{Summary of parameters used in numerical simulations of
    distribution tracking and frequency-map analysis (FMA). Further
    details can be found in Refs.~\cite{md_sim_hel_res_ex_fitterer,
      resexmd2017}.}
  \label{tab:sim_param}
  \begin{ruledtabular}
    \begin{tabular}{lcc}
      Parameter & distribution & FMA \\
      \colrule
      beam &\multicolumn{2}{c}{Beam~1} \\
      beam energy &\multicolumn{2}{c}{450 GeV} \\
      normalized emittance & \q{3.5}{\mu m} & \q{2.5}{\mu m} \\
      $4\sigma$ bunch length & \multicolumn{2}{c}{1.3~ns} \\
      $1\sigma$ bunch length & \multicolumn{2}{c}{9.7~cm} \\
      particle distribution & \parbox{0.4\textwidth}{6D Gaussian
                              distribution with $10^4$ particles}
                               & \parbox{0.4\textwidth}{equally spaced
                                 grid in $x, y$ up to $10\sigma$, with
                                 $(\Delta p / p_0) = 0$} \\
      turns tracked & $10^6$ & $10^4$ \\
      \hline
      optics & \multicolumn{2}{c}{2016 or 2017 injection optics, with
               $\beta^* = \q{11}{m}$ at IP1 and IP5} \\
      machine imperfections & standard errors, with $a_1 = b_1 =
                              0$\footnote{Orbit
                              errors are disabled due to
                              different implementation of the
                              $a_1, b_1$ coefficients in \code{lifetrac} and
                              \code{mad-x}. $b_2$ errors are adjusted to
                              yield an average
                              peak $\beta$-beat of 15\% over 60~seeds, as
                              observed in optics measurements
                              in the LHC.} & no errors \\
      octupoles  & \multicolumn{2}{c}{$I_{\mathrm{MOF}} =
                   \q{+19.6}{A}$, $I_{\mathrm{MOD}} = \q{-19.6}{A}$} \\
      tunes $(Q_x, Q_y)$ & \multicolumn{2}{c}{(64.28, 59.31) for 2016,
                           (62.27, 60.295) for 2017} \\
      chromaticities $(Q'_x, Q'_y)$ & \multicolumn{2}{c}{(+15, +15)} \\
      \hline
      transverse aperture & $5.7\sigma$ &  \\
      longitudinal aperture & $10\sigma$ &  \\
    \end{tabular}
  \end{ruledtabular}
\end{table*}

For the preparation and interpretation of the experiments, two
different types of simulations were performed:
%
\begin{itemize}
\item tracking of a Gaussian distribution, referred to as
  `distribution tracking' in this paper, to obtain particle loss rates
  and emittance evolution;
\item frequency-map analysis (FMA), to visualize the location and
  intensity of resonances~\cite{fmalaskar}.
\end{itemize}
%
For both simulation types, the tracking code
\code{lifetrac}~\cite{lifetrac} was used. The simulation parameters
are summarized in Table~\ref{tab:sim_param}. Further details are given
in Refs.~\cite{md_sim_hel_res_ex_fitterer, resexmd2017}.

@@

The random excitation is a powerful and flexible halo-cleaning
pattern, but it is also potentially harmful for the proton beam
core. The effects of random noise in various machines have been
studied extensively in the past (see, for instance,
Refs.~\cite{noise_lebedev_ssc, noise_alexahin_lhc, noise_2007_ohmi,
  noise_2014_ohmi, md_noise_bbLHC}).

@@

\begin{acknowledgments}
  There are several people we would like to acknowledge for their
  support, insights, participation in studies, and assistance in data
  collection. In particular, Giulia Pa\-potti offered essential
  support for machine operations and helped prepare the 2016
  experiments. Roderik Bruce and Gianluca Va\-len\-ti\-no made
  important suggestions on the design of the experiments. Dmitry
  Sha\-ti\-lov shared his deep understanding of numerical tracking
  simulations with the \code{lifetrac} code. St\'{e}\-phane
  Far\-toukh, Ric\-car\-do De Maria and Ro\-ge\-lio Tom\'{a}s offered
  invaluable assistance in generating the appropiate accelerator
  lattice models. We wish to thank Sergei Na\-gai\-tsev for the
  insightful discussions on the interaction between the transverse
  damping system and resonant excitations. We are grateful to
  En\-ri\-co Bra\-vin, Georges Trad, and the whole beam
  instrumentation team for their crucial support on
  synchrotron-radiation diagnostics. We thank Ste\-fania
  Pa\-pa\-do\-pou\-lou and Fa\-nou\-ria An\-to\-niou for the fruitful
  collaboration on the analysis of the synchrotron-radiation profiles.
